Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-21 _tex/paper2.tex
Built site for gh-pages

To address these issues, this study extends Thurstone's general form
using a systematic and integrated approach based on Causal and Bayesian
inference methods. In addition to improving statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
BTL model assumptions to actively testing whether those assumptions fit
the data under analysis.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key assessment design features relevant
to CJ experiments, such as the selection of judges, stimuli, and
comparisons. Section~\ref{sec-statistical} translates these theoretical
and practical elements into a probabilistic statistical model to analyze
dichotomous pairwise comparison data. Section~\ref{sec-discussion}
discusses the findings, limitations, and challenges and explores avenues
for future research. Finally, Section~\ref{sec-conclusion} summarizes
the study's conclusions.

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of modern CJ data. In
particular, the assumption may not hold when researchers assess complex
traits or heterogeneous stimuli
\citep{Thurstone_1927a, Bramley_2008, Kelly_et_al_2022}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide indications of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}.
These misfit texts may exhibit larger discriminal dispersions or
represent outliers --texts with distinctive characteristics that deviate
markedly from the rest of the sample \citep{Grubbs_1969}. In either
case, comparing misfit texts results in more judgment discrepancies, yet
the BTL model neither accounts for these cases nor offers any means to
address them.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent outlying observations,
the common CJ practice of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
The direction and magnitude of these biases remain unpredictable, as
they depend on which stimuli researchers exclude from the analysis.

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, based on the arguments developed in
Section~\ref{sec-theory-issue1b}, the SCM incorporates the judges'
biases \((B_{kj})\). Together with the outcome, these variables
constitute the preliminary set of endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\). Finally,
the SCM presents the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

\subsection{Limitations and further
research}\label{sec-discussion-limitations}

>>>>

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines Causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity \citep{Perron_et_al_2015}, and
improving statistical accuracy in hypothesis testing, this approach
offers two key advantages. First, it clarifies the interactions among
all actors and processes involved in CJ experiments. Second, it shifts
the current comparative data analysis paradigm from passively accepting
the BTL model assumptions to actively testing whether those assumptions
fit the data under analysis.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key CJ experimental design features,
such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model to analyze dichotomous
pairwise comparison data. Section~\ref{sec-discussion} discusses the
findings, limitations, and challenges and explores avenues for future
research. Finally, Section~\ref{sec-conclusion} summarizes the study's
conclusions.

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of modern CJ data. In
particular, the assumption may not hold when researchers assess complex
traits or heterogeneous stimuli
\citep{Thurstone_1927a, Bramley_2008, Kelly_et_al_2022}, as these traits
and stimuli can introduce judgment discrepancies due to their unique
characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide indications of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citep{Pollitt_2004, Pollitt_2012b, Pollitt_2012a, Goossens_et_al_2018}.
These misfit texts may exhibit larger discriminal dispersions or
represent outliers--texts with distinctive characteristics that deviate
markedly from the rest of the sample \citep{Grubbs_1969}. In either
case, the comparison of these texts results in more judgment
discrepancies, yet the BTL model neither accounts for these cases nor
offers any means to address them, except their elimination from the
analysis.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent texts with different
dispersions or outlying observations, the common CJ practice of
excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
The direction and magnitude of these biases remain unpredictable, as
they depend on which stimuli researchers exclude from the analysis.

@@

In line with Thurstone's theory, SCM~\ref{fig-cj03_scm} depicts the
texts' discriminal processes \((T_{ia}, T_{hb})\) and their discriminal
difference \((D_{iahbjk})\) (see Section~\ref{sec-thurstone_theory}).
Additionally, building on the arguments developed in
Section~\ref{sec-theory-issue1b}, the SCM incorporates a key design
features of CJ experiments: the judges' biases \((B_{kj})\). These
variables, along with the outcome \((O_{iahbjk})\), constitute the
preliminary set of endogenous variables,
\(V = \{ O_{iahbjk}, D_{iahbjk}, T_{ia}, T_{hb}, B_{kj} \}\). Finally,
the SCM presents the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

@@

This study extends Thurstone's general form through a systematic,
integrated approach that combines Causal and Bayesian inference methods.
It begins by formulating the extension as a conceptual model,
represented by an SCM and a DAG. This formulation integrates Thurstone's
core theoretical principles, including the distribution of the
discriminal process, with key CJ experimental design features, such as
the sampling and comparison mechanisms. The study then translates the
SCM into a bespoke statistical model that addresses key limitations of
the BTL model. Specifically, the statistical model accounts for judge
biases, captures the hierarchical structure of stimuli, incorporates
measurement error in trait estimation and hypothesis testing, and
accommodates violations of the equal dispersion assumption. These
improvements potentially enhance measurement reliability and validity
\citep{Perron_et_al_2015} and improve the statistical accuracy of
hypothesis testing.

Beyond these potential benefits, the approach offers two key advantages.
First, it clarifies the interactions among all actors and processes
involved in CJ experiments. Second, it shifts the analytic paradigm from
passively accepting the assumptions of the BTL model to actively testing
their fit with observed data. Together, these advantages create a
principled framework for evaluating best practices in CJ experimental
designs--one that better aligns with the demands of contemporary CJ
assessment contexts \citep{Kelly_et_al_2022} and opens new avenues for
research.

One promising research avenue involves investigating how the
hierarchical structure of stimuli affects CJ experiment designs and
hypotheses testing. One such study could examine whether collecting a
single text per individual provides sufficient information to produce
reliable trait estimates at different levels of the hierarchy. Due to
concerns about the feasibility of the comparison task
\citep{Boonen_et_al_2020}, many CJ studies assess traits using only one
text per individual \citep{citation}. This experimental design
implicitly assumes that within-individual variability is small relative
to between-individual differences--that is, texts produced by the same
individual are relatively homogeneous. However, researchers rarely test
this assumption, even though it directly influences decisions about the
number of texts and participants needed for the experiment, ultimately
affecting the reliability of the trait estimates.

Relatedly, another study could investigate hypotheses about the
characteristics of the texts themselves. Previously, the literature
lacked a direct and integrated method for evaluating these types of
hypotheses. Now, by collecting data on specific stimulus features,
researchers can assess whether these variables produce the expected
effects.
